cardinalR: Generating interesting high-dimensional data structures

### 2025-07-06 paper-cardinalR.Rmd
Updated Treofoil section

>>>>

### Trefoil knots

<!--https://laustep.github.io/stlahblog/posts/TorusKnot4D.html-->

The Trefoil is a closed, nontrivial one-dimensional manifold embedded in $3\text{-}D$ or $4\text{-}D$ space. The trefoil features topological complexity in the form of self-overlaps, making it a valuable test case for evaluating the ability of NLDR methods to preserve global structure, loops, and embeddings in high-dimensional data. Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:trefoil-tb-html)', '\\@ref(tab:trefoil-tb-pdf)'))` summarizes these functions.

```{r}
trefoil_tb <- tibble(fun = c("gen_trefoil4d",
                            "gen_trefoil3d"), 
                        exp = c("Generate a trefoil in $4\\text{-}D$.",
                                "Generate a trefoil in $3\\text{-}D$."))
```

```{r trefoil-tb-html, eval=knitr::is_html_output()}
trefoil_tb |> 
  kable(caption = "cardinalR trefoil data generation functions", col.names = c("Function", "Explanation")) 
```

```{r trefoil-tb-pdf, eval=knitr::is_latex_output()}
trefoil_tb |> 
  kable(caption = "cardinalR trefoil data generation functions", format="latex", col.names = c("Function", "Explanation"), booktabs = T)  |>
  column_spec(1, width = "4cm") |>
  column_spec(2, width = "8cm")
```

#### `gen_trefoil4d()`

The function `gen_trefoil4d()` generates a $4\text{-}D$ trefoil knot using two angular parameters, $\theta$ and $\phi$. The knot is constructed on the $3$-sphere ($S^3 \subset \mathbb{R}^4$), with a band of thickness around the knot path, controlled by `steps`.

The coordinates of the $4\text{-}D$ trefoil knot are defined as $X_1 = \cos(\theta) \cos(\phi)$, $X_2 = \cos(\theta) \sin(\phi)$, $X_3 = \sin(\theta) \cos(1.5 \phi)$, and $X_4 = \sin(\theta) \sin(1.5 \phi)$, where $\theta$ and $\phi$ are angular parameters that trace out the knot's structure on the $3$-sphere $S^3 \subset \mathbb{R}^4$.

The number of `theta` and `phi` values is determined by the `steps` and `n` arguments, respectively. Additional independent noise dimensions can be added via `gen_noisedims()` to extend the structure into $p > 4$ dimensions.

```{r data-trefoil4d, echo=TRUE}
trefoil4d <- gen_trefoil4d(n = 500, p = 4, steps = 5)
```

```{r, eval=FALSE}
langevitour::langevitour(trefoil4d)
```

```{r trefoil4d-proj1}
scaled_data <- scale_data_manual(trefoil4d)

trefoil4d_proj1 <- plot_proj(
  proj_obj = proj_obj1, 
  point_param = c(1.5, 0.2, "#000000"), # size, alpha, color
  plot_limits = c(-0.65, 0.7), 
  title = "a1", 
  cex = 2, 
  axis_text_size = 5,
  is_color = FALSE)

```{r trefoil4d-proj2}

trefoil4d_proj2 <- plot_proj(
  proj_obj = proj_obj2, 
  point_param = c(1.5, 0.2, "#000000"), # size, alpha, color
  plot_limits = c(-0.7, 0.6), 
  title = "a2", 
  cex = 2, 
  axis_text_size = 5,
  is_color = FALSE)

```{r trefoil4d-proj3}

trefoil4d_proj3 <- plot_proj(
  proj_obj = proj_obj3, 
  point_param = c(1.5, 0.2, "#000000"), # size, alpha, color
  plot_limits = c(-0.5, 0.62), 
  title = "a3", 
  cex = 2, 
  axis_text_size = 5,
  is_color = FALSE)

#### `gen_trefoil3d()`

The function `gen_trefoil3d()` generates a $3\text{-}D$ stereographic projection of the $4\text{-}D$ trefoil knot by mapping each point $(X_1, X_2, X_3, X_4) \in \mathbb{R}^4$ to $(X_1', X_2', X_3') \in \mathbb{R}^3$ using the transformation $X_1' = X_1 / (1 - X_4)$, $X_2' = X_2 / (1 - X_4)$, and $X_3' = X_3 / (1 - X_4)$, while excluding points where $X_4 = 1$ to avoid division by zero.

```{r data-trefoil3d, echo=TRUE}
trefoil3d <- gen_trefoil3d(n = 500, p = 4, steps = 5)
```

As with the $4\text{-}D$ version, this function supports optional noise dimensions to embed the projected knot into higher dimensions.

```{r, eval=FALSE}
langevitour::langevitour(trefoil3d)
```

```{r trefoil3d-proj1}
scaled_data <- scale_data_manual(trefoil4d)

trefoil3d_proj1 <- plot_proj(
  proj_obj = proj_obj1, 
  point_param = c(1.5, 0.2, "#000000"), # size, alpha, color
  plot_limits = c(-0.65, 0.7), 
  title = "a1", 
  cex = 2, 
  axis_text_size = 5,
  is_color = FALSE)

```{r trefoil3d-proj2}

trefoil3d_proj2 <- plot_proj(
  proj_obj = proj_obj2, 
  point_param = c(1.5, 0.2, "#000000"), # size, alpha, color
  plot_limits = c(-0.7, 0.6), 
  title = "a2", 
  cex = 2, 
  axis_text_size = 5,
  is_color = FALSE)

```{r trefoil3d-proj3}

trefoil3d_proj3 <- plot_proj(
  proj_obj = proj_obj3, 
  point_param = c(1.5, 0.2, "#000000"), # size, alpha, color
  plot_limits = c(-0.5, 0.62), 
  title = "a3", 
  cex = 2, 
  axis_text_size = 5,
  is_color = FALSE)

```{r, label = "fig-trefoil-proj", fig.cap="Three $2\\text{-}D$ projections from $4\\text{-}D$, for the `trefoil4d` and `trefoil3d` data.", fig.width=15, fig.height=10}

trefoil4d_proj1 + trefoil4d_proj2 + trefoil4d_proj3 + 
  trefoil3d_proj1 + trefoil3d_proj2 + trefoil3d_proj3 +
  plot_layout(ncol = 3, guides = "collect") 
```

### 2025-07-06 paper-cardinalR.tex
Updated Treofoil section

>>>>

\subsubsection{Trefoil knots}\label{trefoil-knots}

The Trefoil is a closed, nontrivial one-dimensional manifold embedded in \(3\text{-}D\) or \(4\text{-}D\) space. The trefoil features topological complexity in the form of self-overlaps, making it a valuable test case for evaluating the ability of NLDR methods to preserve global structure, loops, and embeddings in high-dimensional data. Table \ref{tab:trefoil-tb-pdf} summarizes these functions.

\caption{\label{tab:trefoil-tb-pdf}cardinalR trefoil data generation functions}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}>{\raggedright\arraybackslash}p{8cm}}
\toprule
Function & Explanation\\
\midrule
gen\_trefoil4d & Generate a trefoil in \$4\textbackslash{}text\{-\}D\$.\\
gen\_trefoil3d & Generate a trefoil in \$3\textbackslash{}text\{-\}D\$.\\
\bottomrule
\end{tabular}
\end{table}

\paragraph{\texorpdfstring{\texttt{gen\_trefoil4d()}}{gen\_trefoil4d()}}\label{gen_trefoil4d}

The function \texttt{gen\_trefoil4d()} generates a \(4\text{-}D\) trefoil knot using two angular parameters, \(\theta\) and \(\phi\). The knot is constructed on the \(3\)-sphere (\(S^3 \subset \mathbb{R}^4\)), with a band of thickness around the knot path, controlled by \texttt{steps}.

The coordinates of the \(4\text{-}D\) trefoil knot are defined as \(X_1 = \cos(\theta) \cos(\phi)\), \(X_2 = \cos(\theta) \sin(\phi)\), \(X_3 = \sin(\theta) \cos(1.5 \phi)\), and \(X_4 = \sin(\theta) \sin(1.5 \phi)\), where \(\theta\) and \(\phi\) are angular parameters that trace out the knot's structure on the \(3\)-sphere \(S^3 \subset \mathbb{R}^4\).

The number of \texttt{theta} and \texttt{phi} values is determined by the \texttt{steps} and \texttt{n} arguments, respectively. Additional independent noise dimensions can be added via \texttt{gen\_noisedims()} to extend the structure into \(p > 4\) dimensions.

\begin{verbatim}
trefoil4d <- gen_trefoil4d(n = 500, p = 4, steps = 5)
\end{verbatim}

\paragraph{\texorpdfstring{\texttt{gen\_trefoil3d()}}{gen\_trefoil3d()}}\label{gen_trefoil3d}

The function \texttt{gen\_trefoil3d()} generates a \(3\text{-}D\) stereographic projection of the \(4\text{-}D\) trefoil knot by mapping each point \((X_1, X_2, X_3, X_4) \in \mathbb{R}^4\) to \((X_1', X_2', X_3') \in \mathbb{R}^3\) using the transformation \(X_1' = X_1 / (1 - X_4)\), \(X_2' = X_2 / (1 - X_4)\), and \(X_3' = X_3 / (1 - X_4)\), while excluding points where \(X_4 = 1\) to avoid division by zero.

\begin{verbatim}
trefoil3d <- gen_trefoil3d(n = 500, p = 4, steps = 5)
\end{verbatim}

As with the \(4\text{-}D\) version, this function supports optional noise dimensions to embed the projected knot into higher dimensions.

\begin{figure}[!]
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-trefoil-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `trefoil4d` and `trefoil3d` data.}\label{fig:fig-trefoil-proj}
\end{figure}
